---
title: "BIOPOLE - Chukchi Sea metatranscriptomics analysis report - Manuscript 2"
author: "Nathan"
output-file: index.html
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: true
    code-block-bg: true
editor: source
execute:
  echo: true
  warning: false
  message: false
---
<style>
pre code {
  background-color: #f2f2f2 !important;
  display: block;
  padding: 0.8em;
  border-radius: 6px;
}
code {
  background-color: #f2f2f2 !important;
  padding: 0.15em 0.35em;
  border-radius: 4px;
}
</style>

This report summarises the bioinformatics workflow applied to the Chukchi Sea eRNA samples to explore the metabolic pathways active and highly expressed, invovled in biogeochemical cycles.

## 🧪 Sample Processing, Sequencing & Data Availability

The Centre for Genomic Research (CGR) processed **9 marine eRNA samples** (+ 1 negative control) for metatranscriptomic sequencing as part of project **SSP203072**. The workflow included:

- DNase treatment of all extracted RNA samples.

- rRNA depletion using the *Seawater riboPOOL* kit.

- Construction of **Illumina RNA-seq libraries** from the rRNA-depleted material.

Multiplexing of all libraries and sequencing on one lane of an **Illumina NovaSeq X Plus** (NovaSeq X+) 25B flow cell.

This setup is expected to produce approximately **~325 million read pairs per sample**, providing deep metatranscriptomic coverage suitable for resolving nitrogen-cycling pathways, including denitrification and anammox.

**Funding & administrative details:**

- **SSP ID**: SSP203072

- **Platform**: Illumina NovaSeq X+

- **Funding**: NEOF1819 — Purchase Order P11212-200

All files—including raw data, processing details, and download instructions—are available via:

[Project data download](https://cgr.liv.ac.uk/illum/SSP203072_b7beb589a74e3e61/)

## 📊 Sequencing Output & CGR QC Summary

| Name                                                                 | Raw Reads (M) | Trimmed Reads (M) | Raw Bases (M) | Trimmed Bases (M) |
|----------------------------------------------------------------------|---------------|--------------------|----------------|--------------------|
| 1-S4S                                                                | 422.03        | 418.30             | 126,608.75     | 120,507.35         |
| 2-S4M                                                                | 559.84        | 557.67             | 167,952.06     | 163,451.31         |
| 3-S4B                                                                | 613.03        | 611.42             | 183,908.71     | 179,372.50         |
| <span style="color:red;">4-S13S</span>                               | 9.89          | 7.81               | 2,968.42       | 2,244.11           |
| 5-S13M                                                               | 240.41        | 238.67             | 72,124.26      | 70,520.45          |
| 6-S13B                                                               | 617.78        | 615.94             | 185,334.01     | 181,063.09         |
| 7-S14S                                                               | 526.76        | 524.35             | 158,026.71     | 153,976.49         |
| 8-S14M                                                               | 526.33        | 524.19             | 157,899.00     | 153,074.61         |
| 9-S14B                                                               | 571.17        | 569.59             | 171,349.89     | 166,325.31         |
| <span style="color:red;">10-negative_control</span>                  | 5.92          | 5.16               | 1,776.58       | 1,487.48           |

**Description of Methods**
The raw FASTQ files are trimmed for the presence of Illumina adapter sequences using Cutadapt version 4.5 . The option -O 3 was used, meaning that trimming occurred in reads where at least 3bp of adapter sequence was observed at the 3' end of the read. The reads were further trimmed to remove low quality bases with a minimum window quality score of 20. After trimming, reads shorter than 15 bp were removed. Only read pairs are retained, therefore if a read of a pair is trimmed away then the other read in the pair is also discarded.

::: {.callout-note title="📝 Notes on Low-Yield Samples"}
**4-S13S** returned an exceptionally low number of reads because very little RNA was recovered from the corresponding Sterivex filters. This limited input material resulted in poor library complexity and low overall sequencing yield.

**10-negative_control** is a processing negative control sample and therefore contains no biological RNA. The very low read count is expected and confirms that minimal contamination occurred during library preparation.
:::

# 🧪 Bioinformatics Workflow

## rRNA Removal – **SortMeRNA**

To enrich for mRNA and other non-rRNA transcripts, we removed ribosomal RNA using **SortMeRNA**.

The command below shows the SortMeRNA run that was used for each sample. We used the CGR-trimmed paired-end reads as input and the **smr_v4.3_sensitive_db_rfam_seeds** database for rRNA identification.

```bash
sortmerna \
  --ref "smr_v4.3_sensitive_db_rfam_seeds.fasta" \
  --reads "$R1" \
  --reads "$R2" \
  --threads "$SLURM_CPUS_PER_TASK" \
  --fastx \
  --out2 \
  --paired_in \
  --aligned "${SAMPLE}_rRNA" \
  --other "${SAMPLE}_RNA_clean" \
  --workdir "$WORKDIR"
``` 

For each sample, SortMeRNA produces:

*_rRNA_* files containing reads classified as rRNA

*_RNA_clean_* files containing the rRNA-depleted read pairs, which are carried forward to assembly.

## De Novo Assembly – **MEGAHIT**

We performed a co-assembly of selected rRNA-depleted samples using MEGAHIT, optimised for complex metatranscriptomic data with the meta-sensitive preset. For illustration, the command below shows the co-assembly of samples:

```bash   
megahit 
-1 [station_samples]_RNA_clean_fwd.fq.gz 
-2 [station_samples]_RNA_clean_rev.fq.gz 
-t 60 
--presets meta-sensitive 
--min-contig-len 500 
-m 0.9 
-o St[1:3]_megahit_out
```

Assembly statistics for an example MEGAHIT run (St1) were obtained using seqkit:

```bash
seqkit stats final.contigs.fa
```

For example (St1 assembly):

| Assembly           | # Contigs | Total Length (bp) | Min Length (bp) | Mean Length (bp) | Max Length (bp) |
| ------------------ | --------- | ----------------- | --------------- | ---------------- | --------------- |
| `final.contigs.fa` | 1,432,902 | 1,457,453,119     | 500             | 1,017.1          | 106,426         |

These statistics indicate a highly fragmented but deep assembly, typical of complex marine microbial metatranscriptomes, with contigs ≥500 bp retained for downstream analyses.

## Read Mapping – **Bowtie2** and **Samtools**

To quantify transcript coverage across assembled contigs, we mapped the rRNA-depleted reads back to the MEGAHIT assembly using Bowtie2, followed by BAM processing with Samtools.

First, we built a Bowtie2 index from the fixed assembly:

```bash
bowtie2-build St1_assembly-fixed.fa St1_assembly-fixed
```

Then, for each sample, paired-end reads were aligned and processed as follows:

```bash
IDX="St1_assembly-fixed"
R1="SAMPLE_RNA_clean_fwd.fq.gz"
R2="SAMPLE_RNA_clean_rev.fq.gz"
THREADS=16
OUT="St1_SAMPLE"

Map paired-end reads

bowtie2 -x "${IDX}"
-1 "${R1}"
-2 "${R2}"
-p "${THREADS}"
-S "${OUT}.sam"

Convert to BAM, keep mapped reads only (-F 4), sort and index

samtools view -F 4 -bS "${OUT}.sam" -o "${OUT}-RAW.bam"
samtools sort -@ "${THREADS}" "${OUT}-RAW.bam" -o "${OUT}.bam"
samtools index "${OUT}.bam"
```

The resulting sorted and indexed BAM files (*.bam, *.bam.bai) are used as input for downstream coverage estimation and gene- or contig-level quantification in anvi’o.

## Creating anvi’o Databases

Subsequent steps involved importing the assemblies and mapped reads into anvi’o to facilitate contig profiling, annotation, and interactive exploration of nitrogen-cycle genes (e.g. denitrification and anammox pathways).

In brief, the workflow included:

1. Creating a contigs database from the MEGAHIT assembly

2. Running HMMs and functional annotation (e.g. PFAM, TIGRFAMs, KOfam)

3. Profiling BAM files for each sample to obtain coverage and detection statistics

4. Merging profiles across samples for joint inspection and downstream analyses

Example (commands will vary slightly depending on environment and anvio version):


1. Create contigs database
```bash
anvi-gen-contigs-database -f St1_assembly-fixed.fa -o St1_contigs.db -n "St1 Chukchi Sea metaT"
```
2. Run HMMs (standard anvio collection)
```bash
anvi-run-hmms -c St1_contigs.db --num-threads 20
```
3. Profile each BAM file
```bash
anvi-profile -i St1_SAMPLE.bam -c St1_contigs.db
--output-dir St1_SAMPLE_profile
--num-threads 20
```
4. Merge profiles across samples
```bash
anvi-merge St1_*_profile/PROFILE.db -o St1_merged_profiles -c St1_contigs.db
```

These anvi’o databases form the basis for functional annotation and visualisation of key nitrogen-cycle pathways in the Chukchi Sea metatranscriptomes.

## 🔍 Functional Annotation of genes

To characterise the metabolic potential represented in the metatranscriptomic assembly, we performed extensive functional annotation of the anvi’o contigs database. This step links assembled sequences to known protein families, orthologous groups, and taxonomic markers, enabling downstream interpretation of nitrogen-cycle genes (e.g., denitrification, anammox).

We used the following anvi’o annotation programs:

```bash
anvi-run-hmms
```
Identifies single-copy core genes (SCGs) and key conserved protein families using anvi’o’s curated HMM collections. This provides quality metrics for assemblies and forms the basis for taxonomic inference.

```bash
anvi-run-ncbi-cogs
```
Assigns Clusters of Orthologous Groups (COGs) from NCBI to contigs based on predicted open reading frames. COG categories help identify broad functional roles such as energy production, amino acid metabolism, stress response, etc.
```bash
anvi-run-kegg-kofams
```
Annotates genes using KEGG Orthologs (KOfam HMMs), providing pathway-level information including nitrogen cycling modules (e.g., nitrate reduction, nitrification, denitrification, anammox).
```bash
anvi-run-scg-taxonomy
```
Uses single-copy core gene content to infer taxonomic origin of contigs, giving insight into which microbial groups are contributing to specific nitrogen-cycling pathways.

## 🔬 Metabolic Reconstruction and Pathway Visualisation

Following functional annotation, we performed metabolic inference and pathway-level analysis using anvi’o’s metabolism suite. These tools integrate KEGG Ortholog (KOfam) annotations with curated pathway models to identify which biochemical pathways are represented in the dataset and to visualise how transcripts map onto specific metabolic processes—particularly those involved in nitrogen cycling.

We used the following anvi’o programs:
```bash
anvi-estimate-metabolism
```
Identifies complete and partial KEGG modules across all annotated contigs, providing an overview of metabolic potential (e.g., denitrification, dissimilatory nitrate reduction, anammox).
```bash
anvi-reaction-network
```
Generates reaction-centric metabolic networks based on KEGG reactions detected in the assembly. This allows exploration of how individual reactions connect into broader nitrogen pathways.
```bash
anvi-draw-kegg
```
Renders KEGG metabolic pathway diagrams annotated with detected genes, enabling visual inspection of pathway completeness and highlighting which steps are supported by metatranscriptomic evidence.

These tools collectively provide a pathway-level view of nitrogen cycling activity in the Chukchi Sea microbial community, enabling detailed interpretation of expression patterns associated with denitrification, nitrification, and anammox processes.

## Exploring all transcripts

```bash
anvi-export-gene-coverage-and-detection -c St1_contigs.db
                                        -p St1_MERGED/PROFILE.db 
                                        -O St1_gene_expression
```
We replicate this for the other stations. Then we can import the tables to R.

```{r importing data}
library(tidyverse)

st1 <- read_tsv("data/St1_gene_coverage_annotations.tsv") %>%
  mutate(assembly = "St1")

st2 <- read_tsv("data/St2_gene_coverage_annotations.tsv") %>%
  mutate(assembly = "St2")

st3 <- read_tsv("data/St3_gene_coverage_annotations.tsv") %>%
  mutate(assembly = "St3")

genes_all <- bind_rows(st1, st2, st3)

### Making a tidy table

genes_long <- genes_all %>%
  pivot_longer(
    cols = matches("^St[123]_.*[BMS]$"),
    names_to = "sample",
    values_to = "coverage",
    values_drop_na = TRUE
  ) %>%
  mutate(
    station = str_extract(sample, "^St\\d+"),
    depth_code = str_extract(sample, "[BMS]$"),
    depth = recode(
      depth_code,
      S = "Surface",
      M = "Mid",
      B = "Bottom"
    ),
    KO_ID = na_if(KO_ID, "NA"),
    KO_function = na_if(KO_function, "NA"),
    COG_ID = na_if(COG_ID, "NA"),
    COG_Function = na_if(COG_Function, "NA")
  )

# create a globally unique gene identifier
genes_long <- genes_long %>%
  mutate(
    gene_id = paste(assembly, gene_callers_id, sep = "_")
  )

# Aggregate by KO

ko_long <- genes_long %>%
  filter(!is.na(KO_ID)) %>%
  group_by(
    sample,
    station,
    depth,
    KO_ID,
    KO_function
  ) %>%
  summarise(
    coverage = sum(coverage, na.rm = TRUE),
    n_genes = n_distinct(gene_id),
    .groups = "drop"
  )

# Create a wide KO matrix

ko_matrix <- ko_long %>%
  select(KO_ID, KO_function, sample, coverage) %>%
  pivot_wider(
    names_from = sample,
    values_from = coverage,
    values_fill = 0
  )

# splitting genes with multiple KOs before aggregating

genes_long_ko <- genes_long %>%
  filter(!is.na(KO_ID)) %>%
  separate_longer_delim(
    cols = c(KO_ID, KO_function),
    delim = ";"
  ) %>%
  mutate(
    KO_ID = str_trim(KO_ID),
    KO_function = str_trim(KO_function)
  ) %>%
  filter(KO_ID != "")

# aggregation

ko_long <- genes_long_ko %>%
  group_by(
    sample,
    station,
    depth,
    KO_ID,
    KO_function
  ) %>%
  summarise(
    coverage = sum(coverage, na.rm = TRUE),
    n_genes = n_distinct(gene_id),
    .groups = "drop"
  )

```

```{r}
library(ggplot2)

ggplot(ko_long, aes(coverage)) +
  geom_histogram(bins = 100) +
  scale_x_log10() +
  theme_bw()
```
