---
title: "BIOPOLE - Chukchi Sea metatranscriptomics analysis report - Manuscript 2"
author: "Nathan"
output-file: index.html
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: true
    code-block-bg: true
editor: source
execute:
  echo: true
  warning: false
  message: false
---
<style>
pre code {
  background-color: #f2f2f2 !important;
  display: block;
  padding: 0.8em;
  border-radius: 6px;
}
code {
  background-color: #f2f2f2 !important;
  padding: 0.15em 0.35em;
  border-radius: 4px;
}
</style>

This report summarises the bioinformatics workflow applied to the Chukchi Sea eRNA samples to explore the metabolic pathways active and highly expressed, invovled in biogeochemical cycles.

## 🧪 Sample Processing, Sequencing & Data Availability

The Centre for Genomic Research (CGR) processed **9 marine eRNA samples** (+ 1 negative control) for metatranscriptomic sequencing as part of project **SSP203072**. The workflow included:

- DNase treatment of all extracted RNA samples.

- rRNA depletion using the *Seawater riboPOOL* kit.

- Construction of **Illumina RNA-seq libraries** from the rRNA-depleted material.

Multiplexing of all libraries and sequencing on one lane of an **Illumina NovaSeq X Plus** (NovaSeq X+) 25B flow cell.

This setup is expected to produce approximately **~325 million read pairs per sample**, providing deep metatranscriptomic coverage suitable for resolving nitrogen-cycling pathways, including denitrification and anammox.

**Funding & administrative details:**

- **SSP ID**: SSP203072

- **Platform**: Illumina NovaSeq X+

- **Funding**: NEOF1819 — Purchase Order P11212-200

All files—including raw data, processing details, and download instructions—are available via:

[Project data download](https://cgr.liv.ac.uk/illum/SSP203072_b7beb589a74e3e61/)

## 📊 Sequencing Output & CGR QC Summary

| Name                                                                 | Raw Reads (M) | Trimmed Reads (M) | Raw Bases (M) | Trimmed Bases (M) |
|----------------------------------------------------------------------|---------------|--------------------|----------------|--------------------|
| 1-S4S                                                                | 422.03        | 418.30             | 126,608.75     | 120,507.35         |
| 2-S4M                                                                | 559.84        | 557.67             | 167,952.06     | 163,451.31         |
| 3-S4B                                                                | 613.03        | 611.42             | 183,908.71     | 179,372.50         |
| <span style="color:red;">4-S13S</span>                               | 9.89          | 7.81               | 2,968.42       | 2,244.11           |
| 5-S13M                                                               | 240.41        | 238.67             | 72,124.26      | 70,520.45          |
| 6-S13B                                                               | 617.78        | 615.94             | 185,334.01     | 181,063.09         |
| 7-S14S                                                               | 526.76        | 524.35             | 158,026.71     | 153,976.49         |
| 8-S14M                                                               | 526.33        | 524.19             | 157,899.00     | 153,074.61         |
| 9-S14B                                                               | 571.17        | 569.59             | 171,349.89     | 166,325.31         |
| <span style="color:red;">10-negative_control</span>                  | 5.92          | 5.16               | 1,776.58       | 1,487.48           |

**Description of Methods**
The raw FASTQ files are trimmed for the presence of Illumina adapter sequences using Cutadapt version 4.5 . The option -O 3 was used, meaning that trimming occurred in reads where at least 3bp of adapter sequence was observed at the 3' end of the read. The reads were further trimmed to remove low quality bases with a minimum window quality score of 20. After trimming, reads shorter than 15 bp were removed. Only read pairs are retained, therefore if a read of a pair is trimmed away then the other read in the pair is also discarded.

::: {.callout-note title="📝 Notes on Low-Yield Samples"}
**4-S13S** returned an exceptionally low number of reads because very little RNA was recovered from the corresponding Sterivex filters. This limited input material resulted in poor library complexity and low overall sequencing yield.

**10-negative_control** is a processing negative control sample and therefore contains no biological RNA. The very low read count is expected and confirms that minimal contamination occurred during library preparation.
:::

# 🧪 Bioinformatics Workflow

## rRNA Removal – **SortMeRNA**

To enrich for mRNA and other non-rRNA transcripts, we removed ribosomal RNA using **SortMeRNA**.

The command below shows the SortMeRNA run that was used for each sample. We used the CGR-trimmed paired-end reads as input and the **smr_v4.3_sensitive_db_rfam_seeds** database for rRNA identification.

```bash
sortmerna \
  --ref "smr_v4.3_sensitive_db_rfam_seeds.fasta" \
  --reads "$R1" \
  --reads "$R2" \
  --threads "$SLURM_CPUS_PER_TASK" \
  --fastx \
  --out2 \
  --paired_in \
  --aligned "${SAMPLE}_rRNA" \
  --other "${SAMPLE}_RNA_clean" \
  --workdir "$WORKDIR"
``` 

For each sample, SortMeRNA produces:

*_rRNA_* files containing reads classified as rRNA

*_RNA_clean_* files containing the rRNA-depleted read pairs, which are carried forward to assembly.

## De Novo Assembly – **MEGAHIT**

We performed a co-assembly of selected rRNA-depleted samples using MEGAHIT, optimised for complex metatranscriptomic data with the meta-sensitive preset. For illustration, the command below shows the co-assembly of samples:

```bash   
megahit 
-1 [station_samples]_RNA_clean_fwd.fq.gz 
-2 [station_samples]_RNA_clean_rev.fq.gz 
-t 60 
--presets meta-sensitive 
--min-contig-len 500 
-m 0.9 
-o St[1:3]_megahit_out
```

Assembly statistics for an example MEGAHIT run (St1) were obtained using seqkit:

```bash
seqkit stats final.contigs.fa
```

For example (St1 assembly):

| Assembly           | # Contigs | Total Length (bp) | Min Length (bp) | Mean Length (bp) | Max Length (bp) |
| ------------------ | --------- | ----------------- | --------------- | ---------------- | --------------- |
| `final.contigs.fa` | 1,432,902 | 1,457,453,119     | 500             | 1,017.1          | 106,426         |

These statistics indicate a highly fragmented but deep assembly, typical of complex marine microbial metatranscriptomes, with contigs ≥500 bp retained for downstream analyses.

## Read Mapping – **Bowtie2** and **Samtools**

To quantify transcript coverage across assembled contigs, we mapped the rRNA-depleted reads back to the MEGAHIT assembly using Bowtie2, followed by BAM processing with Samtools.

First, we built a Bowtie2 index from the fixed assembly:

```bash
bowtie2-build St1_assembly-fixed.fa St1_assembly-fixed
```

Then, for each sample, paired-end reads were aligned and processed as follows:

```bash
IDX="St1_assembly-fixed"
R1="SAMPLE_RNA_clean_fwd.fq.gz"
R2="SAMPLE_RNA_clean_rev.fq.gz"
THREADS=16
OUT="St1_SAMPLE"

Map paired-end reads

bowtie2 -x "${IDX}"
-1 "${R1}"
-2 "${R2}"
-p "${THREADS}"
-S "${OUT}.sam"

Convert to BAM, keep mapped reads only (-F 4), sort and index

samtools view -F 4 -bS "${OUT}.sam" -o "${OUT}-RAW.bam"
samtools sort -@ "${THREADS}" "${OUT}-RAW.bam" -o "${OUT}.bam"
samtools index "${OUT}.bam"
```

The resulting sorted and indexed BAM files (*.bam, *.bam.bai) are used as input for downstream coverage estimation and gene- or contig-level quantification in anvi’o.

## Creating anvi’o Databases

Subsequent steps involved importing the assemblies and mapped reads into anvi’o to facilitate contig profiling, annotation, and interactive exploration of nitrogen-cycle genes (e.g. denitrification and anammox pathways).

In brief, the workflow included:

1. Creating a contigs database from the MEGAHIT assembly

2. Running HMMs and functional annotation (e.g. PFAM, TIGRFAMs, KOfam)

3. Profiling BAM files for each sample to obtain coverage and detection statistics

4. Merging profiles across samples for joint inspection and downstream analyses

Example (commands will vary slightly depending on environment and anvio version):


1. Create contigs database
```bash
anvi-gen-contigs-database -f St1_assembly-fixed.fa -o St1_contigs.db -n "St1 Chukchi Sea metaT"
```
2. Run HMMs (standard anvio collection)
```bash
anvi-run-hmms -c St1_contigs.db --num-threads 20
```
3. Profile each BAM file
```bash
anvi-profile -i St1_SAMPLE.bam -c St1_contigs.db
--output-dir St1_SAMPLE_profile
--num-threads 20
```
4. Merge profiles across samples
```bash
anvi-merge St1_*_profile/PROFILE.db -o St1_merged_profiles -c St1_contigs.db
```

These anvi’o databases form the basis for functional annotation and visualisation of key nitrogen-cycle pathways in the Chukchi Sea metatranscriptomes.

## 🔍 Functional Annotation of genes

To characterise the metabolic potential represented in the metatranscriptomic assembly, we performed extensive functional annotation of the anvi’o contigs database. This step links assembled sequences to known protein families, orthologous groups, and taxonomic markers, enabling downstream interpretation of nitrogen-cycle genes (e.g., denitrification, anammox).

We used the following anvi’o annotation programs:

```bash
anvi-run-hmms
```
Identifies single-copy core genes (SCGs) and key conserved protein families using anvi’o’s curated HMM collections. This provides quality metrics for assemblies and forms the basis for taxonomic inference.

```bash
anvi-run-ncbi-cogs
```
Assigns Clusters of Orthologous Groups (COGs) from NCBI to contigs based on predicted open reading frames. COG categories help identify broad functional roles such as energy production, amino acid metabolism, stress response, etc.
```bash
anvi-run-kegg-kofams
```
Annotates genes using KEGG Orthologs (KOfam HMMs), providing pathway-level information including nitrogen cycling modules (e.g., nitrate reduction, nitrification, denitrification, anammox).
```bash
anvi-run-scg-taxonomy
```
Uses single-copy core gene content to infer taxonomic origin of contigs, giving insight into which microbial groups are contributing to specific nitrogen-cycling pathways.

## 🔬 Metabolic Reconstruction and Pathway Visualisation

Following functional annotation, we performed metabolic inference and pathway-level analysis using anvi’o’s metabolism suite. These tools integrate KEGG Ortholog (KOfam) annotations with curated pathway models to identify which biochemical pathways are represented in the dataset and to visualise how transcripts map onto specific metabolic processes—particularly those involved in nitrogen cycling.

We used the following anvi’o programs:
```bash
anvi-estimate-metabolism
```
Identifies complete and partial KEGG modules across all annotated contigs, providing an overview of metabolic potential (e.g., denitrification, dissimilatory nitrate reduction, anammox).
```bash
anvi-reaction-network
```
Generates reaction-centric metabolic networks based on KEGG reactions detected in the assembly. This allows exploration of how individual reactions connect into broader nitrogen pathways.
```bash
anvi-draw-kegg
```
Renders KEGG metabolic pathway diagrams annotated with detected genes, enabling visual inspection of pathway completeness and highlighting which steps are supported by metatranscriptomic evidence.

These tools collectively provide a pathway-level view of nitrogen cycling activity in the Chukchi Sea microbial community, enabling detailed interpretation of expression patterns associated with denitrification, nitrification, and anammox processes.

## Exploring all transcripts

```bash
anvi-export-gene-coverage-and-detection -c St1_contigs.db
                                        -p St1_MERGED/PROFILE.db 
                                        -O St1_gene_expression
```
We replicate this for the other stations. Then we can import the tables to R.

```{r importing data}
library(tidyverse)

st1 <- read_tsv("data/St1_gene_coverage_annotations.tsv") %>%
  mutate(assembly = "St1")

st2 <- read_tsv("data/St2_gene_coverage_annotations.tsv") %>%
  mutate(assembly = "St2")

st3 <- read_tsv("data/St3_gene_coverage_annotations.tsv") %>%
  mutate(assembly = "St3")

genes_all <- bind_rows(st1, st2, st3)

### Making a tidy table

genes_long <- genes_all %>%
  pivot_longer(
    cols = matches("^St[123]_.*[BMS]$"),
    names_to = "sample",
    values_to = "coverage",
    values_drop_na = TRUE
  ) %>%
  mutate(
    station = str_extract(sample, "^St\\d+"),
    depth_code = str_extract(sample, "[BMS]$"),
    depth = recode(
      depth_code,
      S = "Surface",
      M = "Mid",
      B = "Bottom"
    ),
    KO_ID = na_if(KO_ID, "NA"),
    KO_function = na_if(KO_function, "NA"),
    COG_ID = na_if(COG_ID, "NA"),
    COG_Function = na_if(COG_Function, "NA")
  )

# create a globally unique gene identifier
genes_long <- genes_long %>%
  mutate(
    gene_id = paste(assembly, gene_callers_id, sep = "_")
  )

# Aggregate by KO

ko_long <- genes_long %>%
  filter(!is.na(KO_ID)) %>%
  group_by(
    sample,
    station,
    depth,
    KO_ID,
    KO_function
  ) %>%
  summarise(
    coverage = sum(coverage, na.rm = TRUE),
    n_genes = n_distinct(gene_id),
    .groups = "drop"
  )

# Create a wide KO matrix

ko_matrix <- ko_long %>%
  select(KO_ID, KO_function, sample, coverage) %>%
  pivot_wider(
    names_from = sample,
    values_from = coverage,
    values_fill = 0
  )

# splitting genes with multiple KOs before aggregating

genes_long_ko <- genes_long %>%
  filter(!is.na(KO_ID)) %>%
  separate_longer_delim(
    cols = c(KO_ID, KO_function),
    delim = ";"
  ) %>%
  mutate(
    KO_ID = str_trim(KO_ID),
    KO_function = str_trim(KO_function)
  ) %>%
  filter(KO_ID != "")

# aggregation

ko_long <- genes_long_ko %>%
  group_by(
    sample,
    station,
    depth,
    KO_ID,
    KO_function
  ) %>%
  summarise(
    coverage = sum(coverage, na.rm = TRUE),
    n_genes = n_distinct(gene_id),
    .groups = "drop"
  )

```

```{r plot coverage distribution}

ggplot(ko_long, aes(coverage)) +
  geom_histogram(bins = 100) +
  scale_x_log10() +
  theme_bw()
```

We can see that there are some KOs with extremely low coverage (<0.1). It's probably good to filter them out...

### Normalising gene coverages

Gene coverages were normalised using the housekeeping genes **rpoB** and **gyrB**, which encode the β subunit of RNA polymerase and DNA gyrase B, respectively. These conserved core genes were used as reference markers to provide a measure of background cellular transcriptional activity within each sample. For each sample, the geometric mean of rpoB and gyrB coverage was used as the normalisation factor, reducing dependence on the expression of a single reference gene. Target-gene coverage was divided by this reference value, facilitating comparison of relative gene expression across samples and water-column depths while accounting for differences in overall transcript recovery and sequencing effort. As housekeeping-gene expression may vary with community composition and physiological state, the resulting values should be interpreted as relative expression normalised to community-level housekeeping transcription, rather than absolute expression per cell.

```{r normalisation}

# Extracting housekeeping KOs

hk_kos <- c(
  rpoB = "K03043",
  gyrB = "K02470"
)

hk <- ko_long %>%
  filter(KO_ID %in% hk_kos) %>%
  select(sample, KO_ID, coverage) %>%
  pivot_wider(
    names_from = KO_ID,
    values_from = coverage
  ) %>%
  rename(
    rpoB = K03043,
    gyrB = K02470
  )

# Calculate the geometric mean

hk <- hk %>%
  mutate(
    hk_geomean = sqrt(rpoB * gyrB)
  )

# Add the normalisation factor to ko_long

ko_long_norm <- ko_long %>%
  left_join(
    hk %>% select(sample, hk_geomean),
    by = "sample"
  ) %>%
  mutate(
    norm_coverage = coverage / hk_geomean
  )

```

## Exploration of whole-community functional profiles

```{r NMDS, echo=FALSE}

## ============================================================
## Exploration of whole-community functional profiles
## ============================================================

library(tidyverse)
library(vegan)
library(pheatmap)
library(RColorBrewer)

## 1. Sample metadata

sample_meta <- ko_long_norm %>%
  distinct(sample, station, depth)

## 2. Whole-community KO functional profiles using sample-wise relative abundance

ko_long_rel <- ko_long %>%
  group_by(sample) %>%
  mutate(
    rel_abundance = coverage / sum(coverage, na.rm = TRUE)
  ) %>%
  ungroup()

## 3. Create sample x KO matrix for ordination

ko_matrix_rel <- ko_long_rel %>%
  select(KO_ID, sample, rel_abundance) %>%
  group_by(KO_ID, sample) %>%
  summarise(
    rel_abundance = sum(rel_abundance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = KO_ID,
    values_from = rel_abundance,
    values_fill = 0
  )

## Convert to matrix:
## rows = samples
## columns = KOs
## values = relative abundance

mat_rel <- ko_matrix_rel %>%
  column_to_rownames("sample") %>%
  as.matrix()

## 4. Remove uninformative KOs

mat_rel <- mat_rel[
  ,
  apply(mat_rel, 2, var) > 0,
  drop = FALSE
]

## 5. Whole-community ordination

set.seed(1)

nmds <- metaMDS(
  mat_rel,
  distance = "bray",
  k = 2,
  trymax = 100,
  autotransform = FALSE
)

## 6. Extract NMDS sample scores

nmds_scores <- scores(
  nmds,
  display = "sites"
) %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  as_tibble() %>%
  left_join(
    sample_meta,
    by = "sample"
  )

## 7. Plot NMDS

ggplot(
  nmds_scores,
  aes(
    x = NMDS1,
    y = NMDS2,
    color = station,
    shape = depth
  )
) +
  geom_point(size = 4) +
  geom_text(
    aes(label = sample),
    vjust = -1,
    size = 3,
    show.legend = FALSE
  ) +
  theme_minimal(base_size = 13) +
  labs(
    title = "NMDS of KO functional profiles",
    subtitle = paste0(
      "Bray-Curtis dissimilarity; stress = ",
      round(nmds$stress, 3)
    ),
    x = "NMDS1",
    y = "NMDS2",
    color = "Station",
    shape = "Depth"
  )

```

Whole-community transcriptional functional profiles differed among stations and showed consistent vertical structuring, with mid-depth communities displaying distinct functional profiles relative to surface and bottom samples.

Let's look at which KOs are driving these separations:

```{r envfit}
ko_fit <- envfit(nmds, mat_rel, permutations = 0)

ko_vectors <- as.data.frame(
  scores(ko_fit, display = "vectors")
) %>%
  rownames_to_column("KO_ID")

ko_vectors$R2 <- ko_fit$vectors$r

ko_vectors %>%
  arrange(desc(R2)) %>%
  slice_head(n = 20)
```

These top KOs show the functions most strongly associated with the NMDS structure. 
**Note**: because I used permutations = 0, these are not significance-tested. They are simply the KOs whose abundance patterns best match the NMDS geometry.

```{r}
ko_top <- ko_vectors %>%
  arrange(desc(R2)) %>%
  slice_head(n = 20) %>%
  left_join(
    ko_long %>%
      distinct(KO_ID, KO_function),
    by = "KO_ID"
  )

ko_top %>%
  select(KO_ID, KO_function, NMDS1, NMDS2, R2)
```

```{r top_20_KO_abundance}
top20_long <- ko_long_rel %>%
  filter(KO_ID %in% ko_top$KO_ID) %>%
  select(sample, station, depth, KO_ID, KO_function, rel_abundance)

top20_mat <- top20_long %>%
  select(KO_ID, sample, rel_abundance) %>%
  pivot_wider(
    names_from = sample,
    values_from = rel_abundance,
    values_fill = 0
  ) %>%
  column_to_rownames("KO_ID") %>%
  as.matrix()

sample_annotation <- sample_meta %>%
  select(sample, station, depth) %>%
  column_to_rownames("sample")

sample_annotation <- sample_annotation[
  colnames(top20_mat),
  ,
  drop = FALSE
]

pheatmap(
  top20_mat,
  scale = "row",
  cluster_cols = FALSE,
  annotation_col = sample_annotation,
  border_color = NA,
  main = "Top KOs associated with NMDS structure"
)
```
::: {.callout-note title="📝 NMDS interpretation"}

The NMDS revealed clear structuring of overall KO functional profiles among samples (stress = 0.027), with separation associated with both station and depth. The KOs most strongly associated with the ordination included a mixture of core metabolic functions, bacterial transport/envelope functions, and several eukaryotic cellular processes (e.g. translation, cytoskeletal regulation and RNA processing). The heatmap of the top 20 NMDS-associated KOs showed coordinated differences across samples rather than dominance by a single function, with particularly distinct profiles in some bottom and surface samples. These results suggest that the ordination reflects broad shifts in community functional transcriptional composition, potentially influenced by changes in the relative contribution of different taxonomic groups, rather than variation in a single metabolic pathway.

:::

```{r}

## ------------------------------------------------------------
## 4. Functional richness and diversity per sample
##
## Richness = number of detected KOs
## Shannon diversity = diversity of relative KO abundances
## ------------------------------------------------------------

diversity_df <- tibble(
  sample   = rownames(mat_rel),
  richness = rowSums(mat_rel > 0),
  shannon  = diversity(mat_rel, index = "shannon")
) %>%
  left_join(sample_meta, by = "sample")

## KO Shannon diversity

ggplot(diversity_df, aes(depth, shannon, fill = depth)) +
  geom_col() +
  facet_wrap(~station) +
  theme_minimal(base_size = 13) +
  labs(
    title = "KO functional diversity by station and depth",
    y = "Shannon diversity",
    x = NULL
  ) +
  theme(legend.position = "none")
```

KO functional diversity was broadly comparable among stations and depths


```{r N-cycle_heatmap, fig.width=10, fig.height=8}

## 5. Targeted nitrogen-cycle KO heatmap

n_cycle_kos <- tribble(
  ~KO_ID,    ~gene,   ~pathway,

  # Nitrogen fixation
  "K02588",  "nifH",  "Nitrogen fixation",
  "K02586",  "nifD",  "Nitrogen fixation",
  "K02591",  "nifK",  "Nitrogen fixation",

  # Ammonia oxidation
  "K10944",  "amoA",  "Ammonia oxidation",
  "K10945",  "amoB",  "Ammonia oxidation",
  "K10946",  "amoC",  "Ammonia oxidation",
  "K10535",  "hao",   "Hydroxylamine oxidation",

  # Nitrite oxidation
  "K00370",  "nxrA/narG", "Nitrite oxidation / nitrate reduction",
  "K00371",  "nxrB/narH", "Nitrite oxidation / nitrate reduction",

  # Nitrate reduction
  "K02567",  "napA",  "Nitrate reduction",
  "K02568",  "napB",  "Nitrate reduction",

  # Denitrification
  "K00368",  "nirK",  "NO2 -> NO",
  "K15864",  "nirS",  "NO2 -> NO",
  "K04561",  "norB",  "NO -> N2O",
  "K02305",  "norC",  "NO -> N2O",
  "K00376",  "nosZ",  "N2O -> N2",

  # DNRA
  "K03385",  "nrfA",  "DNRA",

  # Anammox
  "K20932",  "hzsA",  "Anammox",
  "K20933",  "hzsB",  "Anammox",
  "K20934",  "hzsC",  "Anammox",
  "K18601",  "hdh",   "Anammox"
)
n_cycle_long <- ko_long_norm %>%
  filter(KO_ID %in% n_cycle_kos$KO_ID) %>%
  left_join(n_cycle_kos, by = "KO_ID")

# Sanity check: flag any expected N-cycle KOs with zero hits
# across all samples, and confirm which ones were actually found
missing_n_cycle_kos <- setdiff(n_cycle_kos$KO_ID, unique(n_cycle_long$KO_ID))
if (length(missing_n_cycle_kos) > 0) {
  message("Not detected in any sample: ",
          paste(missing_n_cycle_kos, collapse = ", "))
}

n_cycle_wide <- n_cycle_long %>%
  mutate(label = paste(gene, KO_ID, sep = " | ")) %>%
  select(label, pathway, sample, norm_coverage) %>%
  pivot_wider(names_from = sample, values_from = norm_coverage, values_fill = 0)

heat_mat <- n_cycle_wide %>%
  select(-label, -pathway) %>%
  as.matrix()
rownames(heat_mat) <- n_cycle_wide$label

annotation_row <- tibble(pathway = n_cycle_wide$pathway) %>%
  as.data.frame()
rownames(annotation_row) <- rownames(heat_mat)

annotation_col <- sample_meta %>%
  filter(sample %in% colnames(heat_mat)) %>%
  column_to_rownames("sample") %>%
  select(station, depth)

pheatmap(
  heat_mat,   # scale up norm_coverage before log1p so small values remain visible
  annotation_row    = annotation_row,
  annotation_col    = annotation_col,
  cluster_rows      = FALSE,
  cluster_cols      = FALSE,
  color             = colorRampPalette(brewer.pal(9, "YlOrRd"))(100),
  main              = "Housekeeping-normalised N-cycle gene coverage",
  fontsize_row      = 9
)

```
The strong expression of **amoABC**, together with high **nirK** and very low **hao**, is consistent with nitrification being dominated by **ammonia-oxidising archaea**, in agreement with previous observations. 
In contrast, genes associated with other nitrogen-cycle pathways were generally expressed at much lower levels across the samples.

```{r C-cycle_heatmap, fig.width=10, fig.height=8}

c_cycle_kos <- tribble(
  ~KO_ID,    ~gene,       ~pathway,

  # ----------------------------------------------------------
  # Calvin-Benson-Bassham cycle
  # ----------------------------------------------------------
  "K01601",  "rbcL",      "Calvin cycle",
  "K01602",  "rbcS",      "Calvin cycle",
  "K00855",  "prk",       "Calvin cycle",

  # ----------------------------------------------------------
  # Reverse TCA cycle
  # ----------------------------------------------------------
  "K15230",  "aclA",      "Reverse TCA cycle",
  "K15231",  "aclB",      "Reverse TCA cycle",

  # ----------------------------------------------------------
  # 3-HP / 4-HB carbon fixation
  # ----------------------------------------------------------
  "K14534",  "abfD",      "3HP/4HB carbon fixation",

  # ----------------------------------------------------------
  # Wood-Ljungdahl pathway
  # ----------------------------------------------------------
  "K14138",  "acsB",      "Wood-Ljungdahl pathway",
  "K00198",  "cooS/acsA", "Wood-Ljungdahl pathway",

  # ----------------------------------------------------------
  # Methanogenesis
  # ----------------------------------------------------------
  "K00399",  "mcrA",      "Methanogenesis",
  "K00401",  "mcrB",      "Methanogenesis",
  "K00402",  "mcrG",      "Methanogenesis",

  # ----------------------------------------------------------
  # Methane oxidation
  # ----------------------------------------------------------
  "K10944",  "pmoA",      "Methane oxidation",
  "K10945",  "pmoB/amoB", "Methane oxidation",
  "K10946",  "pmoC/amoC", "Methane oxidation",

  # Soluble methane monooxygenase
  "K16157",  "mmoX",      "Methane oxidation",
  "K16158",  "mmoY",      "Methane oxidation",
  "K16159",  "mmoZ",      "Methane oxidation",

  # ----------------------------------------------------------
  # Methanol oxidation / methylotrophy
  # ----------------------------------------------------------
  "K14028",  "mxaF",      "Methanol oxidation",
  "K14029",  "mxaI",      "Methanol oxidation",
  "K23995",  "xoxF",      "Methanol oxidation",

  # ----------------------------------------------------------
  # Aerobic carbon monoxide oxidation
  # ----------------------------------------------------------
  "K03518",  "coxS",      "CO oxidation",
  "K03519",  "coxM",      "CO oxidation",
  "K03520",  "coxL",      "CO oxidation"
)
c_cycle_long <- ko_long_norm %>%
  filter(KO_ID %in% c_cycle_kos$KO_ID) %>%
  left_join(c_cycle_kos, by = "KO_ID")

missing_c_cycle_kos <- setdiff(
  c_cycle_kos$KO_ID,
  unique(c_cycle_long$KO_ID)
)

if (length(missing_c_cycle_kos) > 0) {
  message(
    "Not detected in any sample: ",
    paste(missing_c_cycle_kos, collapse = ", ")
  )
}

c_cycle_wide <- c_cycle_long %>%
  mutate(label = paste(gene, KO_ID, sep = " | ")) %>%
  select(label, pathway, sample, norm_coverage) %>%
  pivot_wider(
    names_from = sample,
    values_from = norm_coverage,
    values_fill = 0
  )

c_heat_mat <- c_cycle_wide %>%
  select(-label, -pathway) %>%
  as.matrix()

rownames(c_heat_mat) <- c_cycle_wide$label

c_annotation_row <- tibble(
  pathway = c_cycle_wide$pathway
) %>%
  as.data.frame()

rownames(c_annotation_row) <- rownames(c_heat_mat)

c_annotation_col <- sample_meta %>%
  filter(sample %in% colnames(c_heat_mat)) %>%
  column_to_rownames("sample") %>%
  select(station, depth)

pheatmap(
  c_heat_mat,
  annotation_row = c_annotation_row,
  annotation_col = c_annotation_col,
  cluster_rows = FALSE,
  cluster_cols = TRUE,
  color = colorRampPalette(brewer.pal(9, "YlOrRd"))(100),
  main = "Housekeeping-normalised C-cycle gene coverage",
  fontsize_row = 9
)
```

Carbon-cycle transcription was strongly dominated by Calvin-cycle markers, particularly rbcL and rbcS, indicating that Calvin-cycle carbon fixation was the most prominent autotrophic carbon-assimilation pathway across the samples.

**BUT** rbcL and rbcS have super high values so let's remove them as it could be masking some patterns!

```{r C-cycle_heatmap_2, fig.width=10, fig.height=8}

c_cycle_long <- ko_long_norm %>%
  filter(KO_ID %in% c_cycle_kos$KO_ID,
        !KO_ID %in% c("K01601", "K01602")) %>%
  left_join(c_cycle_kos, by = "KO_ID")

missing_c_cycle_kos <- setdiff(
  c_cycle_kos$KO_ID,
  unique(c_cycle_long$KO_ID)
)

if (length(missing_c_cycle_kos) > 0) {
  message(
    "Not detected in any sample: ",
    paste(missing_c_cycle_kos, collapse = ", ")
  )
}

c_cycle_wide <- c_cycle_long %>%
  mutate(label = paste(gene, KO_ID, sep = " | ")) %>%
  select(label, pathway, sample, norm_coverage) %>%
  pivot_wider(
    names_from = sample,
    values_from = norm_coverage,
    values_fill = 0
  )

c_heat_mat <- c_cycle_wide %>%
  select(-label, -pathway) %>%
  as.matrix()

rownames(c_heat_mat) <- c_cycle_wide$label

c_annotation_row <- tibble(
  pathway = c_cycle_wide$pathway
) %>%
  as.data.frame()

rownames(c_annotation_row) <- rownames(c_heat_mat)

c_annotation_col <- sample_meta %>%
  filter(sample %in% colnames(c_heat_mat)) %>%
  column_to_rownames("sample") %>%
  select(station, depth)

pheatmap(
  c_heat_mat,
  annotation_row = c_annotation_row,
  annotation_col = c_annotation_col,
  cluster_rows = FALSE,
  cluster_cols = TRUE,
  color = colorRampPalette(brewer.pal(9, "YlOrRd"))(100),
  main = "Housekeeping-normalised C-cycle gene coverage",
  fontsize_row = 9
)
```

The main pattern is that the apparent methane oxidation signal (pmoA/B/C) is relatively strong across many samples... **However**, these KOs overlap with amoABC and reflect the archaeal nitrification signal we already identified.

Expression of xoxF suggests active methylotrophy in the water column, potentially reflecting bacterial utilisation of methanol and other C1 compounds released during phytoplankton production and organic-matter degradation.


# unsupervised functional exploration

```{r exploration, fig.width=10, fig.height=10}

## Identify most variable KOs

ko_variable <- ko_long_norm %>%
  group_by(KO_ID, KO_function) %>%
  summarise(
    variance = var(norm_coverage, na.rm = TRUE),
    mean_coverage = mean(norm_coverage, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(mean_coverage > 0) %>%
  arrange(desc(variance))

top_kos <- ko_variable %>%
  slice_head(n = 50)


## Create matrix

top_ko_mat <- ko_long_norm %>%
  filter(KO_ID %in% top_kos$KO_ID) %>%
  select(KO_ID, sample, norm_coverage) %>%
  pivot_wider(
    names_from = sample,
    values_from = norm_coverage,
    values_fill = 0
  ) %>%
  column_to_rownames("KO_ID") %>%
  as.matrix()


## Sample annotation

annotation_col <- sample_meta %>%
  filter(sample %in% colnames(top_ko_mat)) %>%
  column_to_rownames("sample") %>%
  select(station, depth)


## Plot

pheatmap(
  top_ko_mat,
  scale = "row",
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  annotation_col = annotation_col,
  show_rownames = TRUE,
  border_color = NA,
  main = "Most variable KO expression profiles"
)
```

```{r}
ph <- pheatmap(
  top_ko_mat,
  scale = "row",
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  annotation_col = annotation_col,
  show_rownames = TRUE,
  border_color = NA,
  main = "Most variable KO expression profiles", 
  silent = TRUE
)

ko_clusters <- cutree(
  ph$tree_row,
  k = 4
)

ko_cluster_df <- tibble(
  KO_ID = names(ko_clusters),
  cluster = ko_clusters
) %>%
  left_join(
    ko_long_norm %>%
      distinct(KO_ID, KO_function),
    by = "KO_ID"
  ) %>%
  arrange(cluster)

ko_cluster_df
```

```{r}
ko_cluster_summary <- ko_cluster_df %>%
  group_by(cluster) %>%
  summarise(
    n_KOs = n(),
    functions = paste(KO_function, collapse = "; "),
    .groups = "drop"
  )

ko_cluster_summary
```
:: {.callout-note title="📝 Cluster interpretation"}

Clustering of the 50 most variable KOs revealed distinct functional signatures across samples. One cluster was dominated by nutrient-transport and substrate-acquisition genes, suggesting variation in heterotrophic resource utilisation. A small but highly coherent cluster comprising amoC and nirK reflected the previously identified nitrification signal. A larger cluster containing rbcL, rbcS and energy-metabolism genes was consistent with autotrophic and photosynthetic activity, while a fourth cluster contained genes associated with motility, protein-folding stress responses and cytoskeletal functions. Together, these patterns indicate coordinated shifts in major metabolic and cellular strategies across stations and depths.

:::

## Linking function to taxonomy

To get the taxonomy of the genes, we need to go back to Anvi'o and export the gene sequences, classify them with a taxonomic classifier (Kaiju), then import those assignments back into anvi’o.

```bash
anvi-get-sequences-for-gene-calls \
  -c St1_contigs.db \
  --get-aa-sequences \
  -o St1_genes.faa
```

To assign taxonomy to the genes, we will use Kaiju **nr_euk** database

Why: we are working with marine metatranscriptomic genes and our functional clusters clearly contain both prokaryotic and eukaryotic signals. The standard Kaiju RefSeq database is mainly bacteria, archaea and viruses, whereas nr_euk additionally includes fungi and microbial eukaryotes, which is important if we want to capture phytoplankton and other marine eukaryotes.

```bash
kaiju   
  -t ~/ote_db/kaiju_db/nr_euk/nodes.dmp \
  -f ~/ote_db/kaiju_db/nr_euk/kaiju_db_nr_euk.fmi \
  -i St1_genes.faa \
  -o St1_genes_kaiju.out \
  -p \
  -v \
  -z 60
```
2,028,862 query genes were processed

```bash
kaiju-addTaxonNames \
  -t kaiju_db/nr_euk/nodes.dmp \
  -n kaiju_db/nr_euk/names.dmp \
  -i St1_genes_kaiju.out \
  -o St1_genes_kaiju_names.out \
  -r superkingdom,phylum,class,order,family,genus,species
```

Then I can import back the gene-level taxonomy annotations into anvio using:

```bash
anvi-import-taxonomy-for-genes \
  -c St1_contigs.db \
  -i St1_genes_kaiju_names.out \
  -p kaiju
```
For St1, 776,186 splits (54.2%) were annotated taxonomically
For St2, 766,028 splits (41.6%) were annotated taxonomically
For St3, 577,084 splits (44.2%) were annotated taxonomically

However, we don't really need to import the Kaiju output into anvio because we can use it directly and import it to R (because it has already the gene_callers_id and associated taxonomy).
Let's, parse St1_genes_kaiju_names.out in R and join it to genes_long using gene_callers_id (plus assembly to make IDs globally unique).

```{r kaiju_import, eval=FALSE}
read_kaiju_tax <- function(file, assembly) {

  read_lines(file) %>%
    keep(~ str_starts(.x, "C\t")) %>%
    str_split_fixed("\t", 8) %>%
    as_tibble(.name_repair = "minimal") %>%
    set_names(c(
      "status",
      "gene_callers_id",
      "taxon_id",
      "score",
      "match_taxon_ids",
      "accessions",
      "match_sequences",
      "taxonomy"
    )) %>%
    separate(
      taxonomy,
      into = c(
        "superkingdom",
        "phylum",
        "class",
        "order",
        "family",
        "genus",
        "species"
      ),
      sep = ";",
      fill = "right",
      extra = "drop"
    ) %>%
    mutate(
      across(
        c(superkingdom, phylum, class, order, family, genus, species),
        ~ na_if(str_trim(.x), "NA")
      ),
      gene_callers_id = as.integer(gene_callers_id),
      taxon_id = as.integer(taxon_id),
      assembly = assembly,
      gene_id = paste(assembly, gene_callers_id, sep = "_")
    )
}


tax_st1 <- read_kaiju_tax(
  "data/St1_genes_kaiju_names.out",
  "St1"
)

tax_st2 <- read_kaiju_tax(
  "data/St2_genes_kaiju_names.out",
  "St2"
)

tax_st3 <- read_kaiju_tax(
  "data/St3_genes_kaiju_names.out",
  "St3"
)

taxonomy_all <- bind_rows(
  tax_st1,
  tax_st2,
  tax_st3
)

saveRDS(taxonomy_all, "data/taxonomy_all.rds")

```

```{r join_taxonomy}

taxonomy_all <- readRDS("data/taxonomy_all.rds")

# Join it to genes_long

genes_long_tax <- genes_long %>%
  left_join(
    taxonomy_all %>%
      select(
        gene_id,
        taxon_id,
        superkingdom,
        phylum,
        class,
        order,
        family,
        genus,
        species
      ),
    by = "gene_id"
  )

# Check how much taxonomy was recovered among KO-annotated genes

genes_long_tax %>%
  filter(!is.na(KO_ID)) %>%
  distinct(gene_id, .keep_all = TRUE) %>%
  summarise(
    n_KO_genes = n(),
    classified = sum(!is.na(superkingdom)),
    classified_percent = 100 * mean(!is.na(superkingdom))
  )
```

Excellent! 83.4% of my KO-annotated genes have a taxonomic assignment.

That means my taxonomy coverage is very good for the actual functional analysis. I am now comfortable moving on to ask which taxa contribute the expression of the KOs in clusters 1–4.

```{r join_taxo_clusters}

cluster_genes_tax <- genes_long_tax %>%
  filter(!is.na(KO_ID)) %>%
  separate_longer_delim(KO_ID, delim = ";") %>%
  mutate(KO_ID = str_trim(KO_ID)) %>%
  inner_join(
    ko_cluster_df %>% select(KO_ID, cluster),
    by = "KO_ID"
  )

# Summarising taxonomic contribution, for example at genus level:

cluster_tax_genus <- cluster_genes_tax %>%
  filter(!is.na(genus)) %>%
  group_by(cluster, sample, genus) %>%
  summarise(
    coverage = sum(coverage, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(cluster, sample) %>%
  mutate(
    rel_contribution = coverage / sum(coverage)
  ) %>%
  ungroup()

# see the main taxa driving each cluster:

top_cluster_genera <- cluster_genes_tax %>%
  filter(!is.na(genus)) %>%
  group_by(cluster, genus) %>%
  summarise(
    coverage = sum(coverage, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(cluster) %>%
  mutate(
    contribution = coverage / sum(coverage)
  ) %>%
  slice_max(contribution, n = 10) %>%
  arrange(cluster, desc(contribution))

top_cluster_genera
```

```{r taxo_figure, fig.width=8, fig.height=6}
library(tidytext)

cluster_genus_sum <- cluster_genes_tax %>%
  filter(!is.na(genus)) %>%
  group_by(cluster, genus) %>%
  summarise(
    coverage = sum(coverage, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(cluster) %>%
  mutate(
    contribution = coverage / sum(coverage)
  ) %>%
  ungroup()

my_colors <- c(
  "#FF6666", "#8DD3C7", "#377EB8", "#4DAF4A",
  "#984EA3", "#A6D854", "#FFFF33", "#FF7F00",
  "#F781BF", "#999999", "#66C2A5", "#FC8D62",
  "#8DA0CB", "#E78AC3", "#FFD92F", "#E5C494",
  "#B3B3B3", "#A65628", "#FFFFB3", "#BEBADA",
  "#FB8072", "#80B1D3", "#FDB462"
)

# Keep genera >=1% within at least one cluster
cluster_genus_plot <- cluster_genus_sum %>%
  mutate(
    genus_plot = if_else(contribution >= 0.01, genus, "Other")
  ) %>%
  group_by(cluster, genus_plot) %>%
  summarise(
    contribution = sum(contribution),
    .groups = "drop"
  )

# Order genera by total contribution across all clusters
genus_order <- cluster_genus_plot %>%
  filter(genus_plot != "Other") %>%
  group_by(genus_plot) %>%
  summarise(total = sum(contribution), .groups = "drop") %>%
  arrange(desc(total)) %>%
  pull(genus_plot)

cluster_genus_plot <- cluster_genus_plot %>%
  mutate(
    genus_plot = factor(
      genus_plot,
      levels = c(genus_order, "Other")
    )
  )

# Generate enough colours from your palette
genus_colors <- colorRampPalette(my_colors)(
  length(genus_order)
)

names(genus_colors) <- genus_order

genus_colors <- c(
  genus_colors,
  Other = "grey80"
)

ggplot(
  cluster_genus_plot,
  aes(
    x = factor(cluster),
    y = contribution,
    fill = genus_plot
  )
) +
  geom_col(
    width = 0.8,
    color = "white",
    linewidth = 0.2
  ) +
  scale_fill_manual(
    values = genus_colors,
    drop = FALSE
  ) +
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = c(0, 0)
  ) +
  labs(
    title = "Taxonomic contribution to functional KO clusters",
    x = "KO cluster",
    y = "Relative contribution",
    fill = "Genus"
  ) +
  guides(
    fill = guide_legend(
      ncol = 2,
      byrow = TRUE
    )
  ) +
  theme_classic(base_size = 13) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(size = 12),
    legend.text = element_text(size = 9)
  )

```
**Taxonomic contribution to functional KO clusters**. Relative genus-level contribution to the transcript coverage of each of the four KO clusters identified from the 50 most variable KO expression profiles. Genera contributing ≥1% within a cluster are shown individually, while lower-abundance taxa are grouped as Other.


::: {.callout-note title="Figure interpretation"}
The four KO clusters showed distinct taxonomic signatures. Cluster 1 was strongly dominated by *Pelagibacter*, consistent with a resource-acquisition and heterotrophic metabolism profile. Cluster 2 was dominated by nitrifying taxa, particularly *Nitrosopumilus*, with additional contributions from *Nitrosovibrio* and *Nitrosomonas*, supporting the archaeal nitrification signal identified previously. Cluster 3 was taxonomically more diverse and included several phytoplankton genera such as *Chaetoceros*, *Cylindrotheca*, *Micromonas*, *Thalassiosira* and *Phaeocystis*, consistent with autotrophic and photosynthetic functions. Cluster 4 comprised a broader mixture of marine bacterial taxa.
:::